---
title: "Reddic Housing Report"
format:
  html:
    theme: cerulean
    code-fold: true
execute:
  message: false
  warning: false
---

# House Price Estimates

Davin van den Akker

Elijah Kearl

Kimie Jensen

Austin Bitsoi

```{r}
#| include: false
library(tidyverse)
library(pander)
library(ggbeeswarm)
```

```{r}
#| include: false
house <- read_csv("https://raw.githubusercontent.com/byui-cse/cse450-course/master/data/housing.csv")

mini <- read_csv("https://raw.githubusercontent.com/byui-cse/cse450-course/master/data/housing_holdout_test_mini.csv")

house1 <- house %>% 
  
  arrange(price)
```

## Value of our Model

Although direct access to the firm's proprietary valuation algorithms is not available for a point-by-point comparison, we established an objective baseline to quantify model performance. We compared our XGBoost regressor against a neighborhood-median heuristic (median price by zip code) to demonstrate the additional predictive value of our approach.

Commercial valuation workflows are typically more complex than a zip-code median; nevertheless, this baseline highlights the incremental signal captured by our model. As shown in the accompanying performance figure, the XGBoost model reduces prediction error relative to the neighborhood-median baseline across price bands, with the greatest absolute improvements appearing in the core market segments that represent the majority of transactions.


```{r}
error_by_bin <- read_csv("model_error_by_price_bin.csv")

error_wide <- error_by_bin %>%
  select(bin_left, model, rmse) %>%
  pivot_wider(names_from = model, values_from = rmse) %>%
  mutate(
    ymin = pmin(`XGBoost`, `Neighbor median`),
    ymax = pmax(`XGBoost`, `Neighbor median`),
    diff = `Neighbor median` - `XGBoost`
  )

bin_width <- median(diff(error_wide$bin_left), na.rm = TRUE)
estimated_savings <- sum(pmax(error_wide$diff, 0) * bin_width, na.rm = TRUE)
baseline_area <- sum(pmax(error_wide$`Neighbor median`, 0) * bin_width, na.rm = TRUE)
estimated_savings_pct <- ifelse(baseline_area > 0, estimated_savings / baseline_area, NA_real_)
error_wide$fill_label <- "Error reduction area"
area_label_text <- bquote(bold("Percent better vs neighbor median:"~.(scales::percent(estimated_savings_pct, accuracy = 0.1))))

ggplot() +
  geom_ribbon(
    data = error_wide,
    aes(x = bin_left, ymin = ymin, ymax = ymax, fill = fill_label),
    alpha = 0.4
  ) +
  geom_line(data = error_by_bin, aes(x = bin_left, y = rmse, color = model), lwd = 1) +
  geom_point(data = error_by_bin, aes(x = bin_left, y = rmse, color = model), size = 2) +
  labs(
    x = "Actual Price",
    y = "Error margin (RMSE)",
    title = "XGBoost lowers housing pricing estimate error",
    subtitle = "RMSE by price band: XGBoost vs neighbor median"
  ) +
  scale_fill_manual(
    values = c("Error reduction area" = "#d6d6d6"),
    labels = c("Error reduction area" = area_label_text),
    name = ""
  ) +
  scale_color_discrete(name = "") +
  scale_x_continuous(limits = c(0, 5750000), labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.margin = margin(t = 2, r = 0, b = 0, l = 0),
    legend.text = element_text(size = 11)
  ) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.4)))
```


## Data Insights

We evaluated the relationship between sale price and two structural attributes: build quality (`grade`) and observable condition (`condition`).

The first figure displays the distribution of sale price by `grade` (7 denotes average build quality). There is a pronounced positive association between grade and price: prices increase for grades above average, while grades below average exhibit similar median values.

```{r}
ggplot(data = house, mapping = aes(x = as.factor(grade), y = price)) +
  geom_boxplot(fill = "skyblue") +
  #geom_violin() +
  labs(
    x = "House Build Quality",
    y = "House Price (USD)",
    title = "Nicer Building - Higher Price",
    subtitle = "7 is 'average build quality'"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())
```

The second figure summarizes observed prices by condition, using the median sale price within each category to reduce the influence of extreme outliers. This is a critical adjustment; without accounting for these outliers, the skewed data creates the false impression that property condition has no measurable effect on the final sales price.

```{r}
#| echo: false
cond_binned <- house %>%
  filter(!is.na(price)) %>%
  group_by(condition) %>%
  summarise(
    median_price = median(price, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ggplot(cond_binned, aes(x = factor(condition), y = median_price)) +
  geom_point(size = 3, color = '#1f78b4') +
  labs(
    x = 'Condition',
    y = 'Median sale price (USD)',
    title = 'Median sale price by condition',
    subtitle = 'Each point is the median within a condition category'
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw()
```

The price distribution is strongly right-skewed, with relatively few observations above $2M. This imbalance informed our decision to apply a log transformation to price so the model focuses on proportional error rather than being dominated by a small number of extreme transactions.

```{r}
ggplot(house, aes(x = price)) +
  geom_histogram(fill = "#1f78b4", color = "white", bins = 40) +
  labs(
    x = "Price (USD)",
    y = "Frequency of Price",
    title = "Price distribution is right-skewed",
    subtitle = "This skew motivated the log transformation of price"
  ) +
  scale_x_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| include: false
spensive <- house %>% 
  filter(price > 1999999)

cheap <- house %>% 
  filter(price < 2000000)
```


```{r}
#| include: false
ggplot(spensive, aes(x = grade)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(spensive$grade), max(spensive$grade), by = 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())
```

```{r}
#| include: false
ggplot(cheap, aes(x = grade)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(cheap$grade), max(cheap$grade), by = 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())
```


## Technical Details

### Overview and Methodology
We have developed a highly optimized predictive algorithm (an XGBoost regressor) to accurately forecast housing prices. To ensure standard property valuations are not artificially inflated by extreme luxury market outliers, the system applies mathematical adjustments (log-transformations) to the baseline pricing data.

To evaluate a home's true market value, the model processes three core dimensions:

* Property Specifics: Key physical and market attributes, including the effective age of the home (years since renovation), seasonal market timing (month of sale), and the property's size relative to immediate neighbors.

* Spatial and Location Signals: Precise measurements of proximity to downtown hubs, as well as the distance and density of local points of interest.

* Market Comparables: A rigorous nearest-neighbor median price metric. To prevent the model from artificially inflating its own accuracy ("data leakage"), the system is programmed to explicitly exclude a home's own price when evaluating its neighborhood. Furthermore, it calculates these local baselines using only a strictly quarantined set of reference homes, ensuring realistic, unbiased forecasting.

To maximize the system's efficiency, we utilized automated feature selection (RFECV) to systematically strip away noisy data, retaining only the variables that directly drive price.

### Model Selection and Operational Efficiency
As part of our standard due diligence, we benchmarked this approach against several other industry-leading algorithms (including Random Forest, LightGBM, and CatBoost). We also tested a highly complex "ensemble" architecture—running multiple models simultaneously to determine if a combined approach yielded superior results.

Our testing revealed that the multi-model ensemble performed identically to our fully optimized, standalone XGBoost model. Because deploying an ensemble architecture drastically increases computing overhead and long-term maintenance costs for zero added predictive value, we strategically selected the single-model approach.

### Data Validation
Before finalization, the model underwent rigorous cross-validated tuning and was evaluated against a strict holdout set of "unseen" market data. By measuring the average error margin (RMSE) on this withheld data, we have verified that the final standalone model will deliver peak predictive accuracy in real-world scenarios while keeping operational costs and system complexity to a bare minimum.



## Addressing Specific Questions
Question:
What quantifiable evidence do we have that the predictions we come up with are reliable?

Answer:
In an effort for transparency, while testing our model, we made sure to include quantifiable evidence through different choices of metrics. The most non-technical metrics chosen where the following:
- Percent of houses within 5% of their actual price
- Percent of houses within 10% of their actual price
- Percent of houses within 20% of their actual price

More technical oriented metrics include:
- R2 score
- root mean squared error
- median absolute error

Our model scored well on each of these:
- 42% of houses within 5%
- 63% of houses within 10%
- 86% of houses within 20%
- R2 score: 0.88
- RMSE: 100,077
- MAE: 29,140


Question:
Which property types are weighing most heavily in the house prices predicted by your model?

The attributes / property types weighing most heavily on the house prices predicted by our model are the following:

- The Grade of the House
- If there house is on waterfront
- A combined variable of house size, floors, bathrooms and bedrooms
- The median house price of the zipcode 
- The rating of the houses view

Question:
Are there additional factors about these areas that might be affecting prices?

```{r}
house_downtown <- house %>%
  mutate(
    date_year = as.integer(substr(date, 1, 4)),
    years_since_renovation = if_else(yr_renovated == 0, date_year - yr_built, date_year - yr_renovated),
    month = as.integer(substr(date, 5, 6)),
    sqft_living_to_neighbors = sqft_living / sqft_living15
  ) %>%
  filter(!is.na(price))

if (!"distance_to_downtown" %in% names(house_downtown)) {
  haversine_miles <- function(lon1, lat1, lon2, lat2) {
    r <- 3958.8
    to_rad <- pi / 180
    dlon <- (lon2 - lon1) * to_rad
    dlat <- (lat2 - lat1) * to_rad
    a <- sin(dlat / 2)^2 + cos(lat1 * to_rad) * cos(lat2 * to_rad) * sin(dlon / 2)^2
    2 * r * asin(pmin(1, sqrt(a)))
  }

  house_downtown <- house_downtown %>%
    mutate(distance_to_downtown = haversine_miles(long, lat, -122.3321, 47.6062))
}

max_dist <- quantile(house_downtown$distance_to_downtown, 0.95, na.rm = TRUE)
bin_width <- 2

binned <- house_downtown %>%
  filter(distance_to_downtown <= max_dist) %>%
  mutate(dist_bin = floor(distance_to_downtown / bin_width) * bin_width) %>%
  group_by(dist_bin) %>%
  summarise(
    median_price = median(price, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ggplot(binned, aes(x = dist_bin, y = median_price)) +
  geom_line(color = "#1f78b4", linewidth = 1) +
  geom_point(color = "#e31a1c", size = 1.6, alpha = 0.8) +
  labs(
    title = "Median sale price by distance to downtown (binned)",
    subtitle = "Each point is a median within a distance band",
    x = "Distance to downtown (miles)",
    y = "Median sale price (USD)"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw()
```

Answer:
The binned median plot shows that observed sale prices are higher nearer downtown and generally decline as distance increases. This pattern is consistent with stronger demand close to major employment centers and amenities, and it helps explain the elevated prices observed in the Seattle–Bellevue corridor.

## Code 

Our python code can be found here. 

https://github.com/davinnakker/XGboost_project