---
title: "Reddic Housing Report"
format:
  html:
    theme: cerulean
    code-fold: true
execute:
  message: false
  warning: false
---

# House Price Estimates

Davin van den Akker

Elijah Kearl

Kimie Jensen

Austin Bitsoi

```{r}
#| include: false
library(tidyverse)
library(pander)
library(ggbeeswarm)
```

```{r}
#| include: false
house <- read_csv("https://raw.githubusercontent.com/byui-cse/cse450-course/master/data/housing.csv")

mini <- read_csv("https://raw.githubusercontent.com/byui-cse/cse450-course/master/data/housing_holdout_test_mini.csv")

house1 <- house %>% 
  
  arrange(price)
```

## Value of our Model

Although the data we have work with doesn't allow us to directly compare our work with current company practices, we have been able to build a model that is able to guess house prices more accurately than by taking median or mean house prices by zip code. We are not under the impression the company evaluates house price just by median or mean, but we wanted to give you a baseline for our improvements. 

```{r}
ggplot(house1, aes(x = 1:nrow(house1), y = price)) +
  geom_line(color = "blue3", lwd = 1) +
  labs(
    x = "Individual House",
    y = "Price (USD)",
    title = "Model Predictions"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  geom_hline(yintercept = median(house1$price), lwd = .7) +
  #geom_hline(yintercept = mean(house1$price), lwd = .7) +
  annotate(
    "text",
    x = 750,
    y = 750000,
    label = "Median Line"
  ) +
  theme_bw()
```

## Data Insights

An obvious thing to look at is the "niceness" of the house compared to the cost of it. There's two variables worth isolating here: "Grade" and "Condition". 


In the first plot here, we see the distribution of house price by "Grade", or build quality, with 7 being average. The build quality of the house has a strong bearing on the actual price of the house, with anything under "average" being about the same price, and for every step above average, we get more and more price out of the house. 

```{r}
ggplot(data = house, mapping = aes(x = as.factor(grade), y = price)) +
  geom_boxplot(fill = "skyblue") +
  #geom_violin() +
  labs(
    x = "House Build Quality",
    y = "House Price (USD)",
    title = "Nicer Building - Higher Price",
    subtitle = "7 is 'average build quality'"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())
```

The second factor we wanted to look at was the condition of a house, or how nice it currently is. It's surprising that condition has virtually no impact on median price, as the distributions can show. 
```{r}
ggplot(data = house, mapping = aes(x = as.factor(condition), y = price)) +
  geom_boxplot( fill = "skyblue") +
  #geom_violin() +
  labs(
    x = "House Condition",
    y = "House Price (USD)",
    title = "Do we care about house condition?",
    subtitle = "Houses hold consistent median prices regardless of condition."
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())
```

When looking at the distribution of price in our data set, we find that there are hardly any houses over 2 million dollars to actually train our model on.

```{r}
ggplot(house, aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Price (USD)",
    y = "Frequency of Price",
    title = "Houses over $2m are outliers",
    subtitle = "Model Training Difficulty Arises due to This"
  ) +
  scale_x_continuous(labels = scales::dollar) +
  theme_bw()
```

```{r}
#| include: false
spensive <- house %>% 
  filter(price > 1999999)

cheap <- house %>% 
  filter(price < 2000000)
```


```{r}
#| include: false
ggplot(spensive, aes(x = grade)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(spensive$grade), max(spensive$grade), by = 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())
```

```{r}
#| include: false
ggplot(cheap, aes(x = grade)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(cheap$grade), max(cheap$grade), by = 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())
```


## Technical Details











































## Addressing Specific Questions
Question:
What quantifiable evidence do we have that the predictions we come up with are reliable?

Answer:
In an effort for transperency, while testing our model, we made sure to include quantifible evidence through different choices of metrics. The most non-technical metrics chosen where the following:
- Percent of houses within 5% of their actual price
- Percent of houses within 10% of their actual price
- Percent of houses within 20% of their actual price

More technical oriented metrics include:
- R2 score
- root mean squared error
- median absolute error

Our model scored well on each of these:
- 42% of houses within 5%
- 63% of houses within 10%
- 86% of houses within 20%
- R2 score: 0.88
- RMSE: 100,077
- MAE: 29,140


Question:
Which property types are weighing most heavily in the house prices predicted by your model?

The attributes / property types weighing most heavily on the house prices predicted by our model are the following:

- The Grade of the House
- If there house is on waterfront
- A combined variable of house size, floors, bathrooms and bedrooms
- The median house price of the zipcode 
- The rating of the houses view

Question:
Are there additional factors about these areas that might be affecting prices?

```{python}
import polars as pl
import numpy as np
from lets_plot import *
LetsPlot.setup_html()

df = pl.read_csv("clean.csv")
df = (
    df.select(["price", "long", "lat"])
    .filter(pl.col("long") < -121.6)
    .with_columns(np.log10(pl.col("price")))
)

(
    ggplot(data=df, mapping=aes(x='long', y='lat', color='price'))
    + geom_point(size=1.4)
    + scale_color_gradient(low="purple", high="yellow")
    + theme(
        panel_background=element_rect(fill="black"),
        plot_background=element_rect(fill="white"))
    + labs(title="Higher Priced Homes located new 'Eastside' of Seattle",
           subtitle="Home prices sky rocket as you get closer to Bellevue")
)
```

Answer:
  We can see that the closer you get to the big tech hubs and the east side of Seattle the home prices skyrocket especially all around the water.

## Code 

Our python code can be found here. 

https://github.com/davinnakker/XGboost_project