---
title: "Reddic Housing Report"
format:
  html:
    theme: cerulean
    code-fold: true
execute:
  message: false
  warning: false
---

# House Price Estimates

Davin van den Akker

Elijah Kearl

Kimie Jensen

Austin Bitsoi

```{r}
#| include: false
library(tidyverse)
library(pander)
library(ggbeeswarm)
```

```{r}
#| include: false
house <- read_csv("https://raw.githubusercontent.com/byui-cse/cse450-course/master/data/housing.csv")

mini <- read_csv("https://raw.githubusercontent.com/byui-cse/cse450-course/master/data/housing_holdout_test_mini.csv")

house1 <- house %>% 
  
  arrange(price)
```

## Value of our Model

While we do not currently have access to the firm’s proprietary valuation algorithms for a direct side-by-side comparison, it is crucial to establish a quantitative baseline to demonstrate our model's predictive power. To do this, we benchmarked our optimized algorithm against a standard market heuristic: estimating a home's value based purely on its neighborhood's median price.

While we recognize that the firm's current practices are far more sophisticated than a simple median, this baseline proves that our model is successfully capturing complex, multi-dimensional market signals rather than relying on broad generalizations. As illustrated in the accompanying performance chart, our tuned model consistently outperforms the neighborhood median benchmark across all price bands, delivering substantially lower error margins—particularly within the core market segments where the majority of transactions occur.


```{r}
error_by_bin <- read_csv("model_error_by_price_bin.csv")

ggplot(error_by_bin, aes(x = bin_left, y = rmse, color = model)) +
  geom_line(lwd = 1) +
  geom_point(size = 2) +
  labs(
    x = "Actual Price (bin start)",
    y = "RMSE",
    title = "RMSE by Price Band: Tuned vs Fixed XGBoost vs Neighbor Median"
  ) +
  scale_x_continuous(limits = c(0, 7000000), labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw()
```

```{r}
ggplot(house1, aes(x = 1:nrow(house1), y = price)) +
  geom_line(color = "blue3", lwd = 1) +
  labs(
    x = "Individual House",
    y = "Price (USD)",
    title = "Model Predictions"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  geom_hline(yintercept = median(house1$price), lwd = .7) +
  #geom_hline(yintercept = mean(house1$price), lwd = .7) +
  annotate(
    "text",
    x = 750,
    y = 750000,
    label = "Median Line"
  ) +
  theme_bw()
```


## Data Insights

An obvious thing to look at is the "niceness" of the house compared to the cost of it. There's two variables worth isolating here: "Grade" and "Condition". 


In the first plot here, we see the distribution of house price by "Grade", or build quality, with 7 being average. The build quality of the house has a strong bearing on the actual price of the house, with anything under "average" being about the same price, and for every step above average, we get more and more price out of the house. 

```{r}
ggplot(data = house, mapping = aes(x = as.factor(grade), y = price)) +
  geom_boxplot(fill = "skyblue") +
  #geom_violin() +
  labs(
    x = "House Build Quality",
    y = "House Price (USD)",
    title = "Nicer Building - Higher Price",
    subtitle = "7 is 'average build quality'"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())
```

The second factor we wanted to look at was the condition of a house, or how nice it currently is. It's surprising that condition has virtually no impact on median price, as the distributions can show. 
```{r}
ggplot(data = house, mapping = aes(x = as.factor(condition), y = price)) +
  geom_boxplot( fill = "skyblue") +
  #geom_violin() +
  labs(
    x = "House Condition",
    y = "House Price (USD)",
    title = "Do we care about house condition?",
    subtitle = "Houses hold consistent median prices regardless of condition."
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())
```

When looking at the distribution of price in our data set, we find that there are hardly any houses over 2 million dollars to actually train our model on.

```{r}
ggplot(house, aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Price (USD)",
    y = "Frequency of Price",
    title = "Houses over $2m are outliers",
    subtitle = "Model Training Difficulty Arises due to This"
  ) +
  scale_x_continuous(labels = scales::dollar) +
  theme_bw()
```

```{r}
#| include: false
spensive <- house %>% 
  filter(price > 1999999)

cheap <- house %>% 
  filter(price < 2000000)
```


```{r}
#| include: false
ggplot(spensive, aes(x = grade)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(spensive$grade), max(spensive$grade), by = 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())
```

```{r}
#| include: false
ggplot(cheap, aes(x = grade)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(cheap$grade), max(cheap$grade), by = 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())
```


## Technical Details

### Overview and Methodology
We have developed a highly optimized predictive algorithm (an XGBoost regressor) to accurately forecast housing prices. To ensure standard property valuations are not artificially inflated by extreme luxury market outliers, the system applies mathematical adjustments (log-transformations) to the baseline pricing data.

To evaluate a home's true market value, the model processes three core dimensions:

* Property Specifics: Key physical and market attributes, including the effective age of the home (years since renovation), seasonal market timing (month of sale), and the property's size relative to immediate neighbors.

* Spatial and Location Signals: Precise measurements of proximity to downtown hubs, as well as the distance and density of local points of interest.

* Market Comparables: A rigorous nearest-neighbor median price metric. To prevent the model from artificially inflating its own accuracy ("data leakage"), the system is programmed to explicitly exclude a home's own price when evaluating its neighborhood. Furthermore, it calculates these local baselines using only a strictly quarantined set of reference homes, ensuring realistic, unbiased forecasting.

To maximize the system's efficiency, we utilized automated feature selection (RFECV) to systematically strip away noisy data, retaining only the variables that directly drive price.

### Model Selection and Operational Efficiency
As part of our standard due diligence, we benchmarked this approach against several other industry-leading algorithms (including Random Forest, LightGBM, and CatBoost). We also tested a highly complex "ensemble" architecture—running multiple models simultaneously to determine if a combined approach yielded superior results.

Our testing revealed that the multi-model ensemble performed identically to our fully optimized, standalone XGBoost model. Because deploying an ensemble architecture drastically increases computing overhead and long-term maintenance costs for zero added predictive value, we strategically selected the single-model approach.

### Data Validation
Before finalization, the model underwent rigorous cross-validated tuning and was evaluated against a strict holdout set of "unseen" market data. By measuring the average error margin (RMSE) on this withheld data, we have verified that the final standalone model will deliver peak predictive accuracy in real-world scenarios while keeping operational costs and system complexity to a bare minimum.



## Addressing Specific Questions

## Code 

Our python code can be found here. 

https://github.com/davinnakker/XGboost_project